% refs/nnd_mcmc.tex
% Salient algorithmic details from Segert & Wycoff (2025)
% "A Probabilistic Basis for Low-Rank Matrix Learning"  arXiv:2510.05447
% Adapted to the heteroscedastic denoising/completion setting in matlap.

\section*{MCMC Samplers for the Nuclear-Norm Distribution Posterior}

\subsection*{Source paper}
Segert, J.\ \& Wycoff, N.\ (2025).
\emph{A Probabilistic Basis for Low-Rank Matrix Learning}.
arXiv:2510.05447v1.

\subsection*{Model}
Observation model (heteroscedastic; entries missing when $s_{ij}=\infty$):
\[
  y_{ij} \mid x_{ij},\, s_{ij}
  \;\sim\; \mathcal{N}(x_{ij},\, s_{ij}^2), \quad (i,j)\in\Omega.
\]
Nuclear-norm prior with scalar $\lambda>0$:
\[
  p(X \mid \lambda) \;\propto\; \exp\!\bigl(-\lambda\|X\|_*\bigr).
\]
Gamma hyperprior $\lambda\sim\mathrm{Gamma}(a_0,b_0)$ (shape / rate).

\subsection*{Gaussian Scale Mixture (GSM) representation}
The NND prior admits the auxiliary-variable representation
\[
  \exp\!\bigl(-\lambda\|X\|_*\bigr)
  \;=\;
  \min_{Q\succ 0}\;\exp\!\Bigl(-\tfrac{\lambda}{2}\bigl[
    \operatorname{tr}(Q^{-1}\Psi) + \operatorname{tr}(Q)\bigr]\Bigr),
  \quad
  \Psi = X^T X,
\]
achieved at $Q_\star = \Psi^{1/2}$.
Treating $Q$ as an auxiliary variable with the pseudo-joint
\[
  p(X,Q,\lambda)
  \;\propto\;
  \exp\!\Bigl(-\tfrac{\lambda}{2}\bigl[\operatorname{tr}(Q^{-1}\Psi)+\operatorname{tr}(Q)\bigr]\Bigr)
  \cdot p(Y\mid X) \cdot p(\lambda)
\]
gives tractable conditional distributions (see Section~4.2 of the source).


% ===========================================================================
\section*{Sampler 1: Proximal MALA (Sec.\ 4.1, Appendix J.4)}
% ===========================================================================

\paragraph{Algorithm.}
Metropolis-Adjusted Langevin Algorithm using the proximal operator of
$-\log p$ as the proposal mean.  For the heteroscedastic setting, the
proximal operator of $\lambda\|\cdot\|_*$ does not simplify to a closed-form
expression (unlike the homoscedastic case in Appendix~J.3 where the proximal
operator of the negative log-posterior has a closed form via SVT).  We
therefore use a \emph{proximal gradient} step instead (as in the matrix
completion case, Appendix~J.4):

\[
  \bar X = \mathrm{SVT}_{\delta\lambda/2}\!\Bigl(
    X^t + \tfrac{\delta}{2}\, \underbrace{M\odot(Y-X^t)/S^2}_{
      \nabla_X\log p_{\text{lik}}(X^t)}
  \Bigr),
\]
where $M_{ij}=\mathbf{1}[(i,j)\in\Omega]$, $\mathrm{SVT}_\tau(Z)=
U\operatorname{diag}(\max(\sigma_k-\tau,0))V^T$ is singular-value
soft-thresholding, and $\delta>0$ is the proposal scale.

\paragraph{Proposal and accept/reject.}
\[
  X^* \;\sim\; \mathcal{N}(\bar X,\, \delta I),
\]
accepted with probability $\min(1,\alpha)$ where
\[
  \log\alpha
  = \log p(X^*) - \log p(X^t)
  + \log q(X^t\mid X^*) - \log q(X^*\mid X^t),
\]
\[
  \log p(Z)
  = -\tfrac{1}{2}\sum_{(i,j)\in\Omega}\frac{(y_{ij}-z_{ij})^2}{s_{ij}^2}
  - \lambda\|Z\|_*,
\]
\[
  \log q(Z'\mid Z) = -\frac{\|Z'-\bar Z\|_F^2}{2\delta},
\]
with $\bar Z$ the proximal-gradient mean centred at $Z$.

\paragraph{Step size adaptation.}
Robbins--Monro targeting acceptance rate $0.574$ (optimal for proximal MCMC;
Crucinio et al.\ 2023):
\[
  \log\delta_{t+1} = \log\delta_t + c_t(\alpha_t - 0.574),
  \quad c_t = c/\sqrt{t}.
\]

\paragraph{Heteroscedastic adaptation.}
In the homoscedastic case (Appendix~J.3), the proximal operator of the full
negative log-posterior has the closed form
$\mathrm{SVT}_{\delta\lambda(\delta+2\gamma^2)^{-1}}
 \bigl((\delta Y + 2\gamma^2 X^t)(\delta+2\gamma^2)^{-1}\bigr)$.
For heteroscedastic $s_{ij}^2$, no such closed form exists; the proximal
gradient approximation above is used instead, exactly as in the matrix
completion variant of Appendix~J.4.  The MH correction renders the chain
exact regardless of this approximation.


% ===========================================================================
\section*{Sampler 2: GSM Gibbs (Sec.\ 4.2; heteroscedastic adaptation)}
% ===========================================================================

\paragraph{Original SVD Gibbs (homoscedastic case).}
For homoscedastic $s_{ij}^2=\gamma^2$, reparameterise $X=U\operatorname{diag}(\sigma)V^T$
(thin SVD).  The conditional $p(U\mid\sigma,V,Y)$ is matrix von Mises--Fisher
and can be sampled using the Hoff (2009) reflective slice sampler.  For
heteroscedastic noise, this conditional loses its vMF structure because
entry-specific weights $w_{ij}=1/s_{ij}^2$ break the trace-form sufficient
statistic.

\paragraph{GSM Gibbs for heteroscedastic noise.}
We instead Gibbs-sample the auxiliary $Q$ directly.  Define the pseudo-joint
above with $Q$ treated as an auxiliary variable.  Under this joint:

\subparagraph{Step A: Sample $X\mid Q,\lambda,Y,S$.}
Rows are conditionally independent:
$x_i \sim \mathcal{N}(\mu_i,\Sigma_i)$, with
\[
  \Sigma_i^{-1} = \mathrm{diag}(s_{i,\cdot}^{-2}\odot m_{i,\cdot}) + \lambda Q^{-1},
  \qquad
  \mu_i = \Sigma_i\,\mathrm{diag}(s_{i,\cdot}^{-2}\odot m_{i,\cdot})\,y_i^{\text{obs}},
\]
where $m_{i,\cdot}$ is the observation mask for row $i$.
Implemented via Cholesky of the precision matrix.

\subparagraph{Step B: Sample $Q\mid X,\lambda$ (Metropolis-within-Gibbs).}
Eigendecompose $\Psi=X^TX=V\operatorname{diag}(\psi)V^T$.
The conditional on each eigenvalue $d_k$ of $Q$ (in the basis of $V$) is
\[
  p(d_k\mid\psi_k,\lambda)
  \;\propto\; \exp\!\Bigl(-\tfrac{\lambda}{2}\bigl(d_k + \psi_k/d_k\bigr)\Bigr),
  \quad d_k>0.
\]
This is GIG$(1,\lambda,\lambda\psi_k)$ with mode $\sqrt{\psi_k}$ (matching
the CAVI $Q$ update).  We sample via Metropolis with a symmetric log-normal
random walk:
\[
  \log d_k' = \log d_k + \sigma_Q\,\varepsilon,\quad \varepsilon\sim\mathcal{N}(0,1),
\]
\[
  \log\alpha_k
  = -\tfrac{\lambda}{2}(d_k'+\psi_k/d_k'-d_k-\psi_k/d_k)
  + (\log d_k' - \log d_k).
\]
(The second term is the Jacobian in log space; the symmetric log-normal
proposal cancels in the MH ratio.)

\subparagraph{Step C: Sample $\lambda\mid X,Q$.}
Conjugate Gamma update matching the CAVI posterior:
\[
  \lambda \;\sim\; \mathrm{Gamma}(a_0+mn,\; b_0+\operatorname{tr}(Q)).
\]

\paragraph{Posterior mean.}
Return $(1/N)\sum_{t=1}^N X^{(t)}$ over $N$ post-warmup samples.
